Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2022-11-15 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Iterate slides forward. Now just to add some results

* Three-stage Bayesian hierarchical model
\begin{alignat*}{2}
  &\text{(Observations)}     &   \y &\sim p(\y \, | \, \x), \\
  &\text{(Latent field)}     &   \x &\sim p(\x \, | \, \btheta), \\
  &\text{(Hyperparameters)}  &   \qquad \btheta &\sim p(\btheta),
\end{alignat*}
where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_n)$, $\btheta = (\theta_1, \ldots, \theta_m)$
* Interested in learning both $(\btheta, \x)$ from data $\y$
* If the middle layer is Gaussian, then it's a \textcolor{hilit}{latent Gaussian model}
\begin{equation*}
  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
\end{equation*}
* Latent field is typically indexed by spatiotemporal location, such that $n > m$

@@

* `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
* I use "Laplace approximation" to mean approximating the normalising constant with Laplace's method^[Rather than approximating the posterior with a Gaussian, which I call a Gaussian approximation.]
* To get started with `TMB`, write your $f(\x, \btheta)$ in `TMB`'s C++ syntax
* As pseudo-Bayesians, we choose the log-posterior
$$
f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)
$$
* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks as follows

@@

* We can use `TMB` to obtain the Laplace approximation
$$
\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} \label{eq:hypermarginal}
$$
* Integrate out a Gaussian approximation $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ to the latent field
* `TMB` uses automatic differentiation [@griewank2008evaluating] via `CppAD` to do this, as well as help with numerical optimisation routines
* TO-DO: how to describe the approximation for $\x$ with `TMB`

@@

* Integrated nested Laplace approximation (INLA) [@rue2009approximate; @blangiardo2015spatial] is an approach to approximate inference which builds on the Laplace approximation
* Goal is to approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}
* To do so, we require the approximations $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
* There are four steps as to how the method works (bare with me!)

@@

1) First Laplace approximate hyperparameter posterior
\begin{equation}
\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} \label{eq:hypermarginal}
\end{equation}
which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
* Notice that this is the same object we had been working with in `TMB`
* We use this approximation \textcolor{hilit}{nested} within integrals -- hence the name INLA

@@

* Gauss-Hermite quadrature is one way to pick nodes and weights, and is based on the theory of polynomial interpolation
* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) $\implies$ does not require manual parameter tuning!
* Works particularly well when the integrand is pretty Gaussian
* Use $k$ quadrature nodes per dimension, e.g. if $k = 3$ then $3^m$ total nodes

@@

```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ with $3^1 = 3$ nodes. \\label{fig:aghq-1d}")}
knitr::include_graphics("depends/aghq-1d.pdf")
```

```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ with $3^2 = 9$ nodes. \\label{fig:aghq-2d}")}
knitr::include_graphics("depends/aghq-2d.pdf")
```

@@

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
There are more accurate (and complicated) versions which I will talk about in a minute!
\end{tcolorbox}
\end{center}

@@

* Rather than taking the marginals of $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ we could calculate \textcolor{hilit}{a new Laplace approximation} for each $x_i$
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$
* Problem: $n$ can be big, and we will need to recalculate this for each $(x_i, \btheta)$
* So for example, if we use a nested AGHQ strategy with $k_x$ and $k_\theta$ then we will need $k_x k_\theta^m$ recalculations for each $i$

@@

@@

* Notebook for this example at `athowes.github.io/elgm-inf/epil`
* From Section 5.2. of @rue2009approximate previously from BUGS manual
* Patients $i = 1, \ldots, 59$ each either assigned treatment $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$ to help with seizures
* Visits to clinics $j = 1, \ldots, 4$ times with $y_{ij}$ the number of seizures of the $i$th person in the two weeks proceeding their $j$th visit to the clinic
* Covariates age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$

@@

>>>>

* Three-stage Bayesian hierarchical model
\begin{alignat*}{2}
  &\text{(Observations)}     &   \y &\sim p(\y \, | \, \x), \\
  &\text{(Latent field)}     &   \x &\sim p(\x \, | \, \btheta), \\
  &\text{(Hyperparameters)}  &   \qquad \btheta &\sim p(\btheta),
\end{alignat*}
where $\y = (y_1, \ldots, y_n)$, $\x = (x_1, \ldots, x_N)$, $\btheta = (\theta_1, \ldots, \theta_m)$
* Interested in learning both $(\btheta, \x)$ from data $\y$
* If the middle layer is Gaussian, then it's a \textcolor{hilit}{latent Gaussian model}
\begin{equation*}
  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
\end{equation*}
* Latent field is typically indexed by spatiotemporal location, such that $N > m$

@@

* `TMB` [@kristensen2015tmb] is an R package which implements the Laplace approximation for latent variable models
* I use "Laplace approximation" to mean approximating the normalising constant with Laplace's method^[Rather than approximating the posterior with a Gaussian, which I call a Gaussian approximation.]
* To get started with `TMB`, write your $f(\x, \btheta)$ in `TMB`'s C++ syntax
* As pseudo-Bayesians, we choose (something proportional to) the log-posterior
$$
f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)
$$
* For example, for the model $\y \sim \mathcal{N}(\mu, 1)$ with $p(\mu) \propto 1$ then the `TMB` user template looks as follows

@@

* We can use `TMB` to obtain the Laplace approximation
$$
\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} \label{eq:hypermarginal}
$$
* Integrate out a Gaussian approximation $\tilde p_\text{G}(\x \, | \, \btheta, \y)$ to the latent field
* `TMB` uses automatic differentiation [@griewank2008evaluating] via `CppAD` to do this, as well as help with numerical optimisation routines
* We then optimise this to obtain a mode $\hat \btheta$, and a Hessian $\bm{H}$ at the mode

@@

* Integrated nested Laplace approximation (INLA) [@rue2009approximate; @blangiardo2015spatial] is an approach to approximate inference which builds on the Laplace approximation
* Goal is to approximate \textcolor{hilit}{posterior marginals} $\{\tilde p(x_i \, | \, \y)\}_{i = 1}^n$ and $\{\tilde p(\theta_j \, | \, \y)\}_{j = 1}^m$
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, N, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}
* To do so, we require the approximations $\tilde p(\btheta \, | \, \y)$ and $\tilde p(x_i \, | \, \btheta, \y)$
* There are four steps as to how the method works (bare with me!)

@@

1) First Laplace approximate hyperparameter posterior
\begin{equation}
\tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{\tilde p_\text{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} \label{eq:hypermarginal}
\end{equation}
which can be marginalised to get $\tilde p(\theta_j \, | \, \y)$
* Notice that this is the same object we had been working with in `TMB`
* We will use this approximation \textcolor{hilit}{nested} within integrals like this one
$$
\int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) \tilde p_\text{LA}(\btheta \, | \, \y) \text{d}\btheta
$$
hence the name INLA

@@

* Gauss-Hermite quadrature is one way to pick nodes $\z \in \mathcal{Q}(m, k)$ and weights $\omega(\z): \mathcal{Q}(m, k) \to \mathbb{R}$, based on the theory of polynomial interpolation
* The \textcolor{hilit}{adaptive} part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) so that $\btheta(\z) = \hat \btheta + \bm{L}\z$
* Works particularly well when the integrand is pretty Gaussian
* Use $k$ quadrature nodes per dimension, e.g. if $k = 3$ then $3^m$ total nodes

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
Key benefits: no manual tuning, works well (and starting to get some theory) in statistical contexts
\end{tcolorbox}
\end{center}
#

```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ with $3^1 = 3$ nodes. If $k$ is odd then the mode is always included. \\label{fig:aghq-1d}")}
knitr::include_graphics("depends/aghq-1d.pdf")
```

@@

```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ with $3^2 = 9$ nodes. Here we use the product rule so that the points in 2D are just 1D x 1D. \\label{fig:aghq-2d}")}
knitr::include_graphics("depends/aghq-2d.pdf")
```

@@

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
There are more accurate (and complicated) versions which I will talk briefly about in a minute!
\end{tcolorbox}
\end{center}

@@

* Previously had been taking the marginals of $\tilde p_\text{G}(\x \, | \, \btheta, \y)$
* Alternative: calculate \textcolor{hilit}{a new Laplace approximation} for each $x_i$
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{\tilde p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$
* Problem: $N$ can be big, and we will need to recalculate this for each $(x_i, \btheta)$
* Ideas like using $\bm{\mu}(\btheta)_{-i}$ to initialise Newton optimisation to find $\bm{\mu}_{-i}(x_i, \btheta)$ could help

@@

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
Plan: see how long a naive version without these modifications takes, then use this work to get speed-ups as required
\end{tcolorbox}
\end{center}

@@

* Replication of example from Section 5.2. of @rue2009approximate, and previously from BUGS manual
* Patients $i = 1, \ldots, 59$ each either assigned treatment $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$ to help with seizures
* Visits to clinics $j = 1, \ldots, 4$ times with $y_{ij}$ the number of seizures of the $i$th person in the two weeks proceeding their $j$th visit to the clinic
* Covariates age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
Notebook for this example at \texttt{athowes.github.io/elgm-inf/epil}
\end{tcolorbox}
\end{center}

@@

# Thanks for listening!

* Joint work with Alex Stringer (Waterloo) and my supervisors Seth Flaxman (Oxford) and Jeff Eaton (Imperial)
* The code for this project is at `github.com/athowes/elgm-inf`
* You can find me online at `athowes.github.io`
